Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-03-31 paper1_jiturra.Rmd
ya

In summary, we argue that class-embedded social relations play a substantial role in shaping support for redistribution. In this sense, the tendency to form homophilous social ties in terms of social class reinforces attitudes and consolidates opinion similarity. Thus, we hypothesize that segregation in similar class environments polarizes individual class interests so that segregated lower (upper) classes hold more (less) redistributive preferences (H1).

@@

To summarize, the current state of research has led us to establish two competing scenarios based on the reviewed evidence and theoretical assumptions regarding the role of economic inequality in the relationship between class-based segregation and redistributive preferences. On the one hand, if we assume that social relations are eroded in more unequal societies, social classes might be more distant due to lower social trust and participation in formal and informal networks. Therefore, one possible expectation is that in unequal societies, the influence of class segregation on redistribution preferences will be stronger (H2a). In contrast, if we assume that economic inequality has unequal consequences for those in the bottom and top rungs of the class structure in terms of cross-class embeddedness. In that case, it might be possible that the working classes are equally segregated regardless of income inequality compared to the more diverse contact and activity networks of the upper classes. Likewise, in line with evidence that suggests that the upper classes tend to be more concerned about the causes and consequences of economic inequality, our alternative expectation is that inequality will mitigate the role of network segregation on redistribution preferences, particularly among the upper classes (H2b).

@@

**Class-based network segregation**: To create our measure of network homogeneity based on social class, we employed the position generator available in the questionnaire. This instrument has been widely employed in social capital studies, relying on the assumption that access to social resources is based on ties to different positions in the social structure following an ego-centered network approach [@lin_access_1986;@vandergaag_position_2008]. In this case, the position generator shows a list of ten occupations where the respondent declares four possible options through which this tie is classified: "Family or relative", "Close friend", "Someone else I know", and "No one". In the case of the indicator used here, the first three categories are classified as "Knows" = 1 and "Does not know" = 0. On the basis of this, it is possible to obtain the total number of a person's social ties. Subsequently, the occupations were classified into three groups resembling social class positions [@otero_open_2021]. First, the Lawyer, Executive of a large firm, and Human resource manager are classified as high-status. Then, a School teacher, Police officer, and Nurse are considered as intermediate-status. Finally, a car mechanic, bus driver, hairdresser, and home or office cleaner are considered as low-status occupations.

@@

The primary societal characteristic under examination is the current levels of economic inequality, for which we have utilized the Gini Index (post-taxes and transfers) sourced from the World Income Inequality Dataset (WID) [@alvaredo_world_2022]. Additionally, two contextual variables have been incorporated as controls in the multilevel models. Firstly, economic prosperity has been considered to ensure consistency in the estimates for economic inequality, drawing from the World Income Inequality Dataset (WID) [@Schmidt-Catran2016; @finseraas_income_2009]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state has been included [see @edlund_democratic_2015]. This measure, as an index, considers three dimensions: (i) tax revenue as a percentage of gross domestic product (GDP) [@ilo_world_2022]; (ii) social spending as a share of GDP [@ilo_world_2022]; and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear models account for the hierarchical structure of the data, where individuals are nested within countries. Therefore, our first analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model allows us to assess the intraclass correlation, revealing that 13.7% of the variance in redistributive preferences can be attributed to belonging to higher-level units (countries). Subsequently, the micro-level models are estimated to determine the association between network homogeneity as well as the interaction with social class for testing our _segregation_ hypothesis ($H_1$). Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class to determine whether economic inequality moderates the effect of the micro-level interaction ($H_2$). Therefore, we estimate a three-way cross-level interaction to determine if the current levels of income inequality affects the interactions between network homogeneity and social class.  In the macro-level models, all the individual level factors are group-mean centered (CWC) to mitigate possible problems of collinearity [@hox_multilevel_2010]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations  [@hox_multilevel_2010].

@@

Regarding network segregation, Figure \@ref(fig:fig2) depicts the distribution of network homogeneity and social class across countries. In contrast to redistributive preferences, the distribution of network homogeneity is more scattered between countries and social classes. However, it worth to mention at least two interesting findings. First, looking at the sample average (M = 0.36), we can notice that the variation between countries is low (SD = 0.03), Thailand being the society with higher levels of network segregation (M = 0.42), and Iceland the one of the lowest (M = 0.31). Second, it is depicted that the working class network homogeneity drives the pattern of homogeneity. As noticed, in most cases observed, the general pattern is that the working class is highly homogeneous in their networks, except in the Philipines, where the intermediate class shows a slightly higher average homogeneity. In addition, the general pattern is that the service class is less segregated than the intermediate and working class. Besides, an interesting fact is that in some cases (e.g., Mexico), although the working class remains highly segregated (above 0.5), the middle and service classes are similar in their network homogeneity. Overall, one of the most substantial findings is that, without any exception, the working class is more segregated than the services class.

@@

Model 1 integrates our measure of network homogeneity. Initially, as previously noted in the macro associations, individuals with more homogeneous networks tend to exhibit stronger redistributive preferences. This association persists in Model 2, where we control for network size and social trust—two pivotal factors in establishing and consolidating interpersonal networks. However, after including social class into the analysis, the relationship between homogeneity and redistributive preferences appears to reverse, with a reduction in statistical significance (_p_<0.10).

@@

Table \@ref(tab:table2) presents the results of the multilevel models for income inequality. In Model 1, income inequality demonstrates a positive association, as expected, yet it fails to reach significance. Subsequently, Model 2 incorporates economic prosperity through GDP per capita, revealing a negative and significant relationship (p > .05). Furthermore, Model 3 indicates that including the size of the welfare state establishes a positive and significant relationship (_p_ > .05) while GDP per capita maintains its negative association with redistributive preferences (_p_ > .01). In other words, in Model 3, economic prosperity appears to depress redistributive preferences; however, when contextualized within a larger welfare state, greater redistributive preferences are observed.

@@

>>>>

In summary, we argue that class-embedded social relations play a substantial role in shaping support for redistribution. In this sense, the tendency to form homophilous social ties in terms of social class reinforces attitudes and consolidates opinion similarity. Thus, we hypothesize that segregation in similar class environments polarizes individual class interests so that segregated lower (upper) classes hold more (less) redistributive preferences ($H_1$).

@@

To summarize, the current state of research has led us to establish two competing scenarios based on the reviewed evidence and theoretical assumptions regarding the role of economic inequality in the relationship between class-based segregation and redistributive preferences. On the one hand, if we assume that social relations are eroded in more unequal societies, social classes might be more distant due to lower social trust and participation in formal and informal networks. Therefore, one possible expectation is that in unequal societies, the influence of class segregation on redistribution preferences will be stronger ($H_{2a}$). In contrast, if we assume that economic inequality has unequal consequences for those in the bottom and top rungs of the class structure in terms of cross-class embeddedness. In that case, it might be possible that the working classes are equally segregated regardless of income inequality compared to the more diverse contact and activity networks of the upper classes. Likewise, in line with evidence that suggests that the upper classes tend to be more concerned about the causes and consequences of economic inequality, our alternative expectation is that inequality will mitigate the role of network segregation on redistribution preferences, particularly among the upper classes ($H_{2b}$).

@@

**Class-based network segregation**: To create our measure of network homogeneity based on social class, we employed the position generator available in the questionnaire. This instrument has been widely employed in social capital studies, relying on the assumption that access to social resources is based on ties to different positions in the social structure following an ego-centered network approach [@lin_access_1986;@vandergaag_position_2008]. In this case, the position generator shows a list of ten occupations where the respondent declares four possible options through which this tie is classified: "Family or relative", "Close friend", "Someone else I know", and "No one". In the case of the indicator used here, the first three categories are classified as "Knows" = 1 and "Does not know" = 0. Based on this, it is possible to obtain the total number of a person's social ties. Subsequently, the occupations were classified into three groups resembling social class positions [@otero_open_2021]. First, the Lawyer, Executive of a large firm, and Human resource manager are classified as high-status. Then, a School teacher, Police officer, and Nurse are considered as intermediate-status. Finally, a car mechanic, bus driver, hairdresser, and home or office cleaner are considered low-status occupations.

@@

The primary societal characteristic under examination is the current levels of economic inequality, for which we have utilized the Gini Index (post-taxes and transfers) sourced from the World Income Inequality Dataset (WID) [@alvaredo_world_2022]. Additionally, two contextual variables have been incorporated as controls in the multilevel models. Firstly, economic prosperity has been considered to ensure consistency in the estimates for economic inequality, drawing from the World Income Inequality Dataset (WID) [@Schmidt-Catran2016; @finseraas_income_2009]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state has been included [see @edlund_democratic_2015]. This measure, as an index, considers three dimensions: (i) tax revenue as a percentage of gross domestic product (GDP) [@ilo_world_2022]; (ii) welfare generosity as total governmental spending as a share of GDP [@ilo_world_2022]; and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear models account for the hierarchical structure of the data, where individuals are nested within countries. Therefore, our first analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model allows us to assess the intraclass correlation, revealing that 13.7% of the variance in redistributive preferences can be attributed to belonging to higher-level units (countries). Subsequently, the micro-level models are estimated to determine the association between network homogeneity as well as the interaction with social class for testing our _segregation_ hypothesis ($H_1$). Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class to determine whether economic inequality moderates the effect of the micro-level interaction ($H_2$). Therefore, we estimate a three-way cross-level interaction to determine if the current levels of income inequality affect the interactions between network homogeneity and social class.  In the macro-level models, all the individual-level factors are group-mean centered (CWC) to mitigate possible problems of collinearity [@hox_multilevel_2010]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations  [@hox_multilevel_2010].

@@

Regarding network segregation, Figure \@ref(fig:fig2) depicts the distribution of network homogeneity and social class across countries. In contrast to redistributive preferences, the distribution of network homogeneity is more scattered between countries and social classes. However, it is worth to mention at least two interesting findings. First, looking at the sample average (M = 0.36), we can notice that the variation between countries is low (SD = 0.03), with Thailand being the society with higher levels of network segregation (M = 0.42), and Iceland the one of the lowest (M = 0.31). Second, it is depicted that the working class network homogeneity drives the pattern of homogeneity. As noticed, in most cases observed, the general pattern is that the working class is highly homogeneous in their networks, except in the Philipines, where the intermediate class shows a slightly higher average homogeneity. In addition, the general pattern is that the service class is less segregated than the intermediate and working class. Besides, an interesting fact is that in some cases (e.g., Mexico), although the working class remains highly segregated (above 0.5), the middle and service classes are similar in their network homogeneity. Overall, one of the most substantial findings is that, without any exception, the working class is more segregated than the services class.

@@

Model 1 integrates our measure of network homogeneity. Initially, as previously noted in the macro associations, individuals with more homogeneous networks tend to exhibit stronger redistributive preferences. This association persists in Model 2, where we control for network size and social trust—two pivotal factors in establishing and consolidating interpersonal networks. However, after including social class in the analysis, the relationship between homogeneity and redistributive preferences appears to reverse, with a reduction in statistical significance (_p_<0.10).

@@

Table \@ref(tab:table2) presents the results of the multilevel models for income inequality. In Model 1, income inequality demonstrates a positive association, as expected, yet it fails to reach significance. Subsequently, Model 2 incorporates economic prosperity through GDP per capita, revealing a negative and significant relationship (*p* > .05). Furthermore, Model 3 indicates that including the size of the welfare state establishes a positive and significant relationship (_p_ > .05) while GDP per capita maintains its negative association with redistributive preferences (_p_ > .01). In other words, in Model 3, economic prosperity appears to depress redistributive preferences; however, when contextualized within a larger welfare state, greater redistributive preferences are observed.

@@

<!-- _Dicussion_  Luego de la presentación de los resultados, acá se comienza a volver a los argumentos del trabajo y se relaciona cada hipótesis con los resultados. Se sugiere ir por orden, y un párrafo para cada hipótesis. -->

La presente investigación buscó responder dos preguntas centrales. Primero, ¿cuál es el rol de la class-based network segregation en las redistributive preferences? Segundo, ¿juega un rol la desigualdad económica en la relación entre class-based network segregation y las redistributive preferences?

Lo que la evidencia aquí presentada sugiere es que una mayor homogeneidad en términos de clase social tiene un efecto polarizante en las preferencias redistributivas. Así, este estudio ha demostrado que la homogeneidad de la red posee implicancias diferenciadas en las preferencias redistributivas según la posición de clase. Por un lado, para la clase trabajadora, poseer redes sociales más homogéneas se traduce en mayores preferencias redistributivas. En contraste, es posible evidenciar lo opuesto en la clase servicio, en donde redes más homogéneas se traducen en una disminución de las preferencias redistributivas.

En relación a el rol de la desigualdad económica, dos alternativas han sido puestas bajo escrutinio.
Por un lado, nuestra expectativa era que la relación de nivel micro sería _amplificada_ por una mayor desigualdad economica, dado el incremento de la distancia social entre las clases sociales ($H_{2a}$). Por otro lado, hipotetizamos que la desigualdad economica puede _mitigar_ la relación entre segregación y preferenias redistributivias tiende a menguar en tanto aumenta la desigualdad económica ($H_{2b}$). En efecto, lo que encontramos es que en la medida que la desigualdad aumenta la relación a nivel micro se va debilitando, lo cual se traduce en una brecha de clase más acotada, principalmente a través de los cambios en las preferencias redistributivas en la clase servicio.

<!-- _conclusion_  -->

Por un lado, nuestra primera hipótesis se fundamenta en la claimed negative influence of the desigualdad económica en la cohesión social.

Por otro lado, las investigaciones sobre desigualdad y preferencias redistributivas han establecido que las personas en la clase alta tiende a volverse más igualitarias en contextos mayor desigualdad económica [@edlund_democratic_2015;@sachweh_why_2019]. Además, evidencia reciente ha establecido en sociedades más desiguales las personas en posicions más aventajadas tienden a tener mayor diversidad socioeconómica en sus lazos sociales [@otero_differences_2023].

Pueden ser que sus chanches de participar incrementen sus posibilidades de diversidad al tener también mayor confianza - eso hace que también

- network instrument, only a proxy of class positions, mainly the intermediate class positions 
- measurement of redistributive preferences
- causal claims and endogeneity problems
